---
title: "Vector Autoregressive Models"
author: "Tyler J. Brough"
date: "November 21, 2016"
output: html_document 
---

## Vector Autoregressive Models

Consider the following two equations:

$$
\begin{aligned}
y_{1,t} &= a y_{1,t-1} + b y_{2,t-1} + v_{1,t} \\
y_{2,t} &= c y_{1,t-1} + d y_{2,t-1} + v_{2,t} 
\end{aligned}
$$

These two equations comprise a vector autoregression (VAR). A VAR is the extension of the autoregressive (AR) model to
the case in which there is more than one variable under study. The VAR above is a bivariate VAR(1) (lag-one). Generally, a VAR can consist of $K$ variables and have $p$ lags. 

In a VAR model each variable is treated as endogenous. There is one equation for each variable in the system, and each
equation consists of lags of it's own variable plus each of the others. Technically speaking, this is called a *reduced
form* VAR, which we will distinguish from a *structural VAR* in just a bit. 

We can write the VAR(p) model more generally as:

$$
\begin{aligned}
y_{1,t} &= \alpha_{1} + \sum\limits_{i=1}^{p} a_{1i} y_{1,t-i} + \sum\limits_{i=1}^{p} b_{1i} y_{2,t-i} + \cdots \sum\limits_{i=1}^{p} c_{1i} y_{k,t-i} + v_{1,t} \\
y_{2,t} &= \alpha_{2} + \sum\limits_{i=1}^{p} a_{2i} y_{1,t-i} + \sum\limits_{i=1}^{p} b_{2i} y_{2,t-i} + \cdots \sum\limits_{i=1}^{p} c_{2i} y_{k,t-i} + v_{2,t} \\
\vdots \\
y_{k,t} &= \alpha_{k} + \sum\limits_{i=1}^{p} a_{ki} y_{1,t-i} + \sum\limits_{i=1}^{p} b_{ki} y_{2,t-i} + \cdots \sum\limits_{i=1}^{p} c_{ki} y_{k,t-i} + v_{k,t} 
\end{aligned}
$$

We can also write the VAR model in matrix form:

$$
Y_{t} = \Gamma + A_{1} Y_{t-1} + A_{2} Y_{t-2} + \cdots + A_{p} Y_{t-p} + V_{t}
$$

Where:

$$
Y_{t} =
 \begin{bmatrix} 
  y_{1,t} \\
  y_{2,t} \\
  \vdots  \\
  y_{k,t} 
 \end{bmatrix}
$$

$$
A_{i} = 
 \begin{bmatrix}
  a_{i,11} & a_{i,12} & \cdots & a_{i,1k} \\
  a_{i,21} & a_{i,22} & \cdots & a_{i,2k} \\
  \vdots                                  \\
  a_{i,k1} & a_{i,k2} & \cdots & a_{i,kk} 
 \end{bmatrix}
$$

$$
V_{t} = 
 \begin{bmatrix}
  v_{1,t} \\
  v_{2,t} \\
  \vdots  \\
  v_{k,t}
 \end{bmatrix}
$$



### Typical Uses of VAR Models

VARs are the one of the most important econometric tools in empirical macroeconomics and finance. They have three main
uses:

1. Forecasting
2. Impulse Response Functions
3. Forecast Error Variance Decomposition

We will focus on using the VAR for calculating Impulse Response Functions (IRFs).

Please see [Stock and Watson](http://www.jstor.org/stable/2696519?seq=1#page_scan_tab_contents) for a very nice overview of VAR modeling. 

### VAR Estimation 

It turns out that the VAR model, at least as we have presented it here, is a special case of a more general model called
a Seemingly Unrelated Regressions (SUR) Model. The VAR is a SUR model for which the right-hand side variables are all
the same in each equation. When that is the case there is no added benefit from systems estimation, so we can estimate
the VAR model by applying OLS regression equation-by-equation. 

An example of estimating a VAR model in R will be given below.


### Impulse Response Functions

In applied work, it is often of interest to know the _response_ of one variable to _impulse_ in another variable in a system
containing a number of further variables. If there is a reaction in the one variable due to an impulse in the other
variable one might call the latter causal for the former. This kind of analysis is sometimes called _dynamic multiplier
analysis_. 

Imagine we are given the following three-variable lag-three VAR model:

$$
\begin{bmatrix} y_{1,t} \\ y_{2,t} \\ y_{3,t} \end{bmatrix} = \begin{bmatrix} .5 & 0 & 0 \\ .1 & .1 & .3 \\ 0 & .2 & .3  \end{bmatrix} \begin{bmatrix} y_{1,t-1} \\ y_{2,t-1} \\ y_{3,t-1}
\end{bmatrix} + \begin{bmatrix} u_{1,t} \\ u_{2,t} \\ u_{3,t}   \end{bmatrix}
$$

___Note:___ Here we are not worrying about estimation, and we are assuming we know the "true" form of the $A_{1}$ matrix. 

Tracing out a unit shock in the first variable in period $t = 0$ in this system we get

$$
y_{0} = \begin{bmatrix} y_{1,0} \\ y_{2,0} \\ y_{3,0} \end{bmatrix} = \begin{bmatrix} u_{1,0} \\ u_{2,0} \\ u_{3,0}   \end{bmatrix} = \begin{bmatrix} 1 \\ 0 \\ 0  \end{bmatrix}
$$

$$
y_{1} = \begin{bmatrix} y_{1,1} \\ y_{2,1} \\ y_{3,1} \end{bmatrix} = A_{1} y_{0} = \begin{bmatrix} .5 \\ .1 \\ 0  \end{bmatrix}
$$

$$
y_{2} = \begin{bmatrix} y_{1,2} \\ y_{2,2} \\ y_{3,2} \end{bmatrix} = A_{1} y_{1} = A_{1}^{2} y_{0} = \begin{bmatrix} .25 \\ .06 \\ .02  \end{bmatrix}
$$

Continuing in this manner, it turns out that $y_{i} = (y_{1,i}, y_{2,i}, y_{3,i})^{\prime}$ is just the first column of $A_{1}^{i}$. It turns out that a unit shock in $y_{2,t}$ ($y_{3,t}$) at $t = 0$ after $i$ periods, results in a vector $y_{i}$ which is just the second (third) column of $A_{1}^{i}$. Thus, the elements of $A_{1}^{i}$ represent the effects of the unit shocks in the variables of the system after $i$ periods. Therefore, they are called impulse responses or dynamic multipliers. 

This is the simplest way to see how impulse responses work, but usually they are calculated from the Vector Moving Average form of the VAR model. Just as with a univariate $AR(p)$ model, a $VAR(p)$ model can be inverted to an infinite-order $VMA(\infty)$ model. The coeficients of the VMA model are the dynamic multipliers. 


### VAR Modeling in R

```{r}
suppressMessages(require(vars))
```

We can look at the use of VAR models in R with the `vars` package and its builtin data set on the Canadian macroeconomy. 

```{r}
data(Canada)
```

First let's use the AIC to specify a proper number of lags, choosing $10$ as the maximum possible lag:

```{r}
VARselect(Canada, lag.max = 10, type = "const")
```

The AIC selects a $p = 3$ order model, which we can now fit as follows:

```{r}
fit <- VAR(Canada, p = 3, type = "const")
summary(fit)
```

As you can see the `summary` function gives a lot of output information. We can see the details of the dataset as follows:

```{r}
help(Canada)
```

Let's look at the effects of a unit shock in the `prod` variable to the `e` variable by calculating the impulse response function out just a few steps for presentation purposes:

```{r}
irf(fit, impulse = "prod", response = "e", n.ahead = 10)
```

We can also plot the IRF:

```{r}
irf.e <- irf(fit, impulse = "prod", response = "e", n.ahead = 100)
plot(irf.e)
```

As you can see, R makes VAR modeling simple. The key trick is understanding what you are doing!




