---
title: "The Natural Exponential Function"
author: "Tyler J. Brough"
date: "October 17, 2016"
output: html_document
---

## Introduction

This note is a quick introduction to the _natural exponential function_. 

See the following wikipedia articles for more details:

- <https://en.wikipedia.org/wiki/Exponential_function>

- <https://en.wikipedia.org/wiki/E_(mathematical_constant)>


## Exponential Functions

Exponential functions take the following form:

$$
f(x) = b^{x}
$$

in which the parameter $b$ is called the base and the variable $x$ occurs as an exponent. 

One particularly helpful exponential function is called the _natural exponential function_. 

## Brief Review of Compounding Relations


Recall from your introductory finance courses that we use the following formula to calculate periodic compounding:

$$
\left(1 + \frac{r}{n} \right)^{n \times T}
$$

We get more and more frequent compounding as $n$ increases for some interest rate $r$ and time horizon $T$. What happens as $n \rightarrow \infty$? We can use calculus to solve this problem. 

But first, let's just use a simple computer experiment to investigate what happens. Let's look at a simplified version of the above formula:

$$
\left(1 + \frac{1}{x} \right) ^{x}
$$

For this investigation let's first set up a user-defined function in `R` for the compounding function:

```{r}
f <- function(x)
{
  result <- (1 + (1/x))^x
  return(result)
}
```

Now, let's use the function for larger and larger values of $x$. Let's increase by powers of 10 to speed things up:

```{r}
powers <- 1:8
base <- 10

for(i in powers)
{
  value <- f(base^i)
  print(value)
}
```

From this computer experiment we can see that the values converge to a certain value, which we will denote as $e \approx 2.718282$. The number $e$ is called ___Euler's number___ for the mathematician Euler who first discovered it. 

This is a simple way to intuitively check the result from calculus for continuous compounding:

$$
\lim_{n \rightarrow \infty} \left(1 + \frac{r}{n} \right)^{n \times T} = e^{r \times T}
$$


## What is Natural about the Natural Exponential Function?

So what is _natural_ about the _natural exponential function_? It's that as we let the value of $x$ go to infinity we get the exponential function with $e$ as it's base. It's thus a natural choice for the base, and we call it the natural exponential function. For finance this is indeed a natural choice as it gives continuous compounding. 

It is also mathematically convenient to calculate the future value of a continuouslly compounded rate, $r$ and maturity $T$ as:

$$
FV = e^{r \times T}
$$

where $T$ is the maturity of the cash flows (expressed in annual terms, i.e. one year is $T = 1$).

We can also get present values that are continuouslly compounded as follows:


$$
PV = e ^ {-r \times T}
$$


